Exploring the Validity Range of PLL Analytical Models

### 2025-05-16 CaseStudies/Example3_2_LimitsoftheTheorySoFar/Report/article/article.qmd
Including bode diagrams for the PLL of Example 3.2.

def format(value):
    abs_value = abs(value)
    if abs_value >= 1e9:
        result = f"{value / 1e9:.2f}\\,\\text{{G}}"
    elif abs_value >= 1e6:
        result = f"{value / 1e6:.2f}\\,\\text{{M}}"
    elif abs_value >= 1e3:
        result = f"{value / 1e3:.2f}\\,\\text{{k}}"
    elif abs_value >= 1:
        result = f"{value * 1e0:.2f}\\,"
    elif abs_value >= 1e-3:
        result = f"{value * 1e3:.2f}\\,\\text{{m}}"
    elif abs_value >= 1e-6:
        result = f"{value * 1e6:.2f}\\,\\mu"
    elif abs_value >= 1e-9:
        result = f"{value * 1e9:.2f}\\,\\text{{n}}"
    elif abs_value >= 1e-12:
        result = f"{value * 1e12:.2f}\\,\\text{{p}}"
    elif abs_value >= 1e-15:
        result = f"{value * 1e15:.2f}\\,\\text{{f}}"
    else:
        result = f"{value:.0e}\\,"

    return Markdown(f"${result}$")
```

@@

To compute the natural frequency of the loop, the following equation is used:

$$
\begin{split}
\omega_{3\text{dB}}&=\omega_n\sqrt{1+2\zeta^2+\sqrt{4\zeta^4+4\zeta^2+2}}\implies\\
\omega_n&=\frac{\omega_{3\text{dB}}}{\sqrt{1+2\zeta^2+\sqrt{4\zeta^4+4\zeta^2+2}}}
\end{split}
$$

@@

# Parâmetros iniciais
N = 5
A0 = 1
KVCO = 10e6
I = 1e-3
Kphase = I / (2 * np.pi)
K = A0 * Kphase * KVCO / N
```

@@

# Análise de resposta em frequência
omega_n_meas = np.sqrt(KVCO * Kphase / (N * (C1 + C2)))
zeta_meas = (C1 * R1 / 2) * np.sqrt(KVCO * Kphase / (N * (C1 + C2)))
t_m_meas=1/(omega_n_meas*math.sqrt(1-zeta_meas**2))*math.atan(math.sqrt(1-zeta_meas**2)/zeta_meas)
theta_e_m_n_meas = math.sin(omega_n_meas*math.sqrt(1-zeta_meas**2)*t_m_meas)/(math.sqrt(1-zeta_meas**2))*math.e**(-zeta_meas*omega_n_meas*t_m_meas)
theta_e_m_meas = 2 * math.pi

@@

# Criar vetor da entrada (degrau de amplitude 5)
step_amplitude = 2*step_m_ang_meas
u = np.ones_like(t) * step_amplitude

@@

# Gerar gráfico
plt.figure(figsize=(8, 5))
plt.plot(t_out * 1e6, y_out, label='Resposta ao Degrau')
plt.plot(t_out * 1e6, u, linestyle=':', color='black', label='Entrada Degrau')
plt.axhline(final_value + tol, color='red', linestyle='--', linewidth=0.8, label='Faixa ±1%')
plt.axhline(final_value - tol, color='red', linestyle='--', linewidth=0.8)
plt.axvline(ts_graphical * 1e6, color='green', linestyle='--', linewidth=1.2, label=f'Settling time (gráfico): {ts_graphical*1e6:.2f} µs')
plt.xlabel('Tempo [µs]')
plt.ylabel('Resposta')
plt.title('Resposta ao Degrau com Tempo de Acomodação (99%)')
plt.grid(True)
plt.legend()
plt.tight_layout()
plt.show()
```

@@

These component values define the loop filter required to achieve the desired dynamic behavior of the PLL.

@@

With the second-order PLL properly defined, the transfer function of this system is given by:

@@

import sympy as sp
from IPython.display import display, Markdown

# Define variáveis simbólicas
s, wn, z = sp.symbols('s omega_n_meas zeta_meas')

# Define numerador e denominador simbolicamente
Hnum = 2 * z * wn * s + wn**2
Hden = s**2 + 2 * z * wn * s + wn**2

H = Hnum / Hden

H_subs = H.subs({wn: omega_n_meas, z: zeta_meas}).evalf(4)

# Converte para LaTeX
latex_str = sp.latex(H_subs)

# Exibe como Markdown numerado
display(Markdown(f"$$H(s) = {latex_str}$$ {{#eq-transfer-function}}"))

@@

A step with the same amplitude as the previously calculated frequency step was also applied in phase, that is, $\Delta\theta=$ `{python} format_with_pi_latex(step_m_ang_meas)`rad/s. Applying this signal to [Eq. @eq-transfer-function], the response of this system to the phase step is obtained. With this step, the settling time achieved is $t_s=$ `{python} format(t_s_meas)`s. Through simulation, it is possible to verify that the obtained value is $t_s=$ `{python} format(ts_graphical)` s. The simulated response can be visualized in [Fig. @fig-pll-step-response].

>>>>

def format_graph(value):
    abs_value = abs(value)
    if abs_value >= 1e9:
        result = f"{value / 1e9:.2f} G"
    elif abs_value >= 1e6:
        result = f"{value / 1e6:.2f} M"
    elif abs_value >= 1e3:
        result = f"{value / 1e3:.2f} k"
    elif abs_value >= 1:
        result = f"{value * 1e0:.2f} "
    elif abs_value >= 1e-3:
        result = f"{value * 1e3:.2f} m"
    elif abs_value >= 1e-6:
        result = f"{value * 1e6:.2f} μ"
    elif abs_value >= 1e-9:
        result = f"{value * 1e9:.2f} n"
    elif abs_value >= 1e-12:
        result = f"{value * 1e12:.2f} p"
    elif abs_value >= 1e-15:
        result = f"{value * 1e15:.2f} f"
    else:
        result = f"{value:.0e} "

    return result
```

def format_no_markdown(value):
    abs_value = abs(value)
    if abs_value >= 1e9:
        result = f"{value / 1e9:.2f}\\,\\text{{G}}"
    elif abs_value >= 1e6:
        result = f"{value / 1e6:.2f}\\,\\text{{M}}"
    elif abs_value >= 1e3:
        result = f"{value / 1e3:.2f}\\,\\text{{k}}"
    elif abs_value >= 1:
        result = f"{value * 1e0:.2f}\\,"
    elif abs_value >= 1e-3:
        result = f"{value * 1e3:.2f}\\,\\text{{m}}"
    elif abs_value >= 1e-6:
        result = f"{value * 1e6:.2f}\\,\\mu"
    elif abs_value >= 1e-9:
        result = f"{value * 1e9:.2f}\\,\\text{{n}}"
    elif abs_value >= 1e-12:
        result = f"{value * 1e12:.2f}\\,\\text{{p}}"
    elif abs_value >= 1e-15:
        result = f"{value * 1e15:.2f}\\,\\text{{f}}"
    else:
        result = f"{value:.0e}\\,"

    return result
```

def format(value):
    result = format_no_markdown(value)
    return Markdown(f"${result}$")
```

@@

To compute the frequency bandwidth of the loop, the following equation is used:

$$
\omega_{3\text{dB}}=\omega_n\sqrt{1+2\zeta^2+\sqrt{4\zeta^4+4\zeta^2+2}}
$$ {#eq-frequency-bandwidth}

Rearanging the [Eq. @eq-frequency-bandwidth], it's possible to compute the natural frequency in terms of frequency bandwidth:

$$
\omega_n=\frac{\omega_{3\text{dB}}}{\sqrt{1+2\zeta^2+\sqrt{4\zeta^4+4\zeta^2+2}}}
$$ {#eq-natural-frequency-from-bandwidth}

@@

# Parâmetros iniciais
N = 5
A0 = 1
KVCO = 10e3
I = 1e-3
Kphase = I / (2 * np.pi)
K = A0 * Kphase * KVCO / N
```

@@

# Análise de resposta em frequência
omega_n_meas = np.sqrt(KVCO * Kphase / (N * (C1 + C2)))
f_n_meas = omega_n_meas/2/np.pi
zeta_meas = (C1 * R1 / 2) * np.sqrt(KVCO * Kphase / (N * (C1 + C2)))
t_m_meas=1/(omega_n_meas*math.sqrt(1-zeta_meas**2))*math.atan(math.sqrt(1-zeta_meas**2)/zeta_meas)
theta_e_m_n_meas = math.sin(omega_n_meas*math.sqrt(1-zeta_meas**2)*t_m_meas)/(math.sqrt(1-zeta_meas**2))*math.e**(-zeta_meas*omega_n_meas*t_m_meas)
theta_e_m_meas = 2 * math.pi

omega_bw_meas = omega_n_meas * np.sqrt(1+2*zeta**2+np.sqrt(4*zeta**4+4*zeta**2+2))
f_bw_meas = omega_bw_meas/2/np.pi

@@

# Criar vetor da entrada (degrau de amplitude 5)
step_amplitude = step_m_ang_meas
u = np.ones_like(t) * step_amplitude

@@

# Gerar gráfico
plt.figure(figsize=(8, 5))
plt.plot(t_out * 1e6, y_out, label='Resposta ao Degrau')
plt.plot(t_out * 1e6, u, linestyle=':', color='black', label='Entrada Degrau')
plt.axhline(final_value + tol, color='red', linestyle='--', linewidth=0.8, label='Faixa ±1%')
plt.axhline(final_value - tol, color='red', linestyle='--', linewidth=0.8)
plt.axvline(ts_graphical * 1e6, color='green', linestyle='--', linewidth=1.2, label=f'Settling time (gráfico): {ts_graphical*1e6:.2f} µs')
plt.xlabel('Time [µs]')
plt.ylabel('Phase [rad/s]')
plt.title('Resposta ao Degrau com Tempo de Acomodação (99%)')
plt.grid(True)
plt.legend()
plt.tight_layout()
plt.show()
```

@@

These component values define the loop filter required to achieve the desired dynamic behavior of the PLL. With these components, it's possible to extract another set of the system parameters, i.e., $f_n=\omega_n/2\pi$, $\zeta$ and $f_\mathrm{bw}=\omega_\mathrm{bw}/2\pi$. The values are $f_n=$ `{python} format(f_n_meas)`Hz, $\zeta=$ `{python} round(zeta_meas, 3)` and $f_\mathrm{bw}=$ `{python} format(f_bw_meas)`Hz.

@@

where $C_s$ is the paralel between $C_1$ and $C_2$. With the PLL values properly defined, the transfer function of this system is given by:

@@

from scipy import signal
from IPython.display import display, Markdown

def format_eng(x):
    if x == 0:
        return "0"
    exponent = int("{:.0e}".format(x).split('e')[1])
    base = x / (10**exponent)
    return f"{base:.2f}e{exponent}"

def poly_to_str(coeffs):
    terms = []
    degree = len(coeffs) - 1
    for i, coef in enumerate(coeffs):
        power = degree - i
        if abs(coef) < 1e-12:  # ignora coeficientes muito pequenos
            continue
        term = ""
        coef_str = format_eng(coef)
        if not (abs(coef - 1) < 1e-12 and power != 0):
            term += coef_str
        if power > 0:
            term += "s"
            if power > 1:
                term += f"^{power}"
        terms.append(term)
    return " + ".join(terms)

# Em scipy.signal, num e den são arrays dentro de listas (mesmo para SISO)
num = H_sys.num[0][0] if isinstance(H_sys.num[0], (list, tuple)) else H_sys.num
den = H_sys.den[0][0] if isinstance(H_sys.den[0], (list, tuple)) else H_sys.den

num_str = poly_to_str(num)
den_str = poly_to_str(den)

# Exibe como equação no Jupyter
display(Markdown(f"$$H(s) = \\frac{{{num_str}}}{{{den_str}}}$$ {{#eq-transfer-function}}"))

@@

A step with the same amplitude as the previously calculated frequency step was also applied in phase, that is, $\Delta\theta=$ `{python} format_with_pi_latex(step_m_ang_meas)`rad/s. Applying this signal to [Eq. @eq-transfer-function], the response of this system to the phase step is obtained. With this step, the settling time achieved is $t_s=$ `{python} format(t_s_meas)`s. Through simulation, it is possible to verify that the obtained value is $t_s=$ `{python} format(ts_graphical)`s. The simulated response can be visualized in [Fig. @fig-pll-step-response].

# Gerar a resposta em frequência
w, mag, phase = signal.bode(H_sys)

# Encontrar o ponto onde a magnitude cruza -3 dB
cutoff_index = np.argmax(mag <= -3)
cutoff_freq_ang = w[cutoff_index]
cutoff_freq_lin = cutoff_freq_ang/2/np.pi
cutoff_mag = mag[cutoff_index]

# Plotando o diagrama de Bode
plt.figure(figsize=(10, 6))

# Magnitude
plt.subplot(2, 1, 1)
plt.semilogx(w, mag, label='Magnitude (dB)')
plt.axvline(x=cutoff_freq_ang, color='r', linestyle='--', label='Frequência de corte (-3 dB)')
plt.axhline(y=-3, color='gray', linestyle=':', label='-3 dB')
plt.scatter(cutoff_freq_ang, cutoff_mag, color='red')  # Marcar o ponto

# Adiciona o valor da frequência no gráfico
plt.text(cutoff_freq_ang, cutoff_mag - 5, f'{format_graph(cutoff_freq_lin)}Hz', color='red',
         ha='center', va='top', fontsize=9, bbox=dict(facecolor='white', edgecolor='red', boxstyle='round'))

plt.title('Resposta em Frequência - Diagrama de Bode')
plt.ylabel('Magnitude (dB)')
plt.grid(True, which="both", ls="--")
plt.legend()

# Fase
plt.subplot(2, 1, 2)
plt.semilogx(w, phase)
plt.axvline(x=cutoff_freq_ang, color='r', linestyle='--')
plt.ylabel('Phase (graus)')
plt.xlabel('Frequency (rad/s)')
plt.grid(True, which="both", ls="--")

plt.tight_layout()
plt.show()
```

## TODO: Analyze control voltage to verify if cycle slipping occurs if $\Delta\omega>\Delta\omega_m$
